---
title: "Algorithm Reference: the 37 Binning Engines"
author: "José Evandeilton Lopes"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Algorithm Reference: the 37 Binning Engines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
options(width = 100, digits = 4)
```

```{r libs}
library(OptimalBinningWoE)
```

## What this covers

The package ships **28 algorithms** covering **37 algorithm/feature-type
combinations**: 21 work on numerical features, 16 on categorical ones, and 9
handle both. This vignette is the reference for choosing among them and for
knowing what each one actually does — including where an algorithm's name does
not describe its implementation, which is true more often than one would like.

Every algorithm is also documented individually on its own help page. What you
get here that you do not get there is the comparison: coverage, families,
measured cost, and the handful of cross-cutting behaviours that only become
visible when the engines are placed side by side.

```{r registry}
alg <- obwoe_algorithms()
c(algorithms = nrow(alg),
  numerical = sum(alg$numerical),
  categorical = sum(alg$categorical),
  both = sum(alg$numerical & alg$categorical),
  multinomial = sum(alg$multinomial))
```

## Two ways to run an algorithm

**Through `obwoe()`**, which detects feature types, dispatches, and returns a
model object that the rest of the package understands:

```{r via-obwoe}
data_path <- system.file("extdata", "germancredit.csv.gz",
                         package = "OptimalBinningWoE")
gc_data <- utils::read.csv(gzfile(data_path), stringsAsFactors = FALSE)
gc_data$target <- as.integer(1L - gc_data$credit_risk)
gc_data$credit_risk <- NULL

fit <- obwoe(gc_data, target = "target", feature = c("duration", "purpose"),
             algorithm = "jedi", min_bins = 2, max_bins = 5)
fit$summary[, c("feature", "type", "algorithm", "n_bins", "total_iv")]
```

**Or by calling the engine directly**, which returns a plain list for a single
feature and no model object:

```{r via-wrapper}
direct <- ob_numerical_jedi(target = gc_data$target,
                            feature = gc_data$duration,
                            min_bins = 2, max_bins = 5)
data.frame(bin = direct$bin, count = direct$count,
           woe = round(direct$woe, 4), iv = round(direct$iv, 4))
```

### `obwoe()` does not forward algorithm-specific parameters

This is the one thing worth knowing before reading the rest. `obwoe()` builds a
fixed argument list — `min_bins`, `max_bins`, `bin_cutoff`, `max_n_prebins`,
`convergence_threshold`, `max_iterations`, and `bin_separator` for categorical
features — and passes nothing else. An algorithm's own parameters are **not**
reachable through it, and passing them via `control.obwoe()` fails **silently**:

```{r forwarding}
via_obwoe <- obwoe(gc_data, target = "target", feature = "duration",
                   algorithm = "dmiv",
                   control = control.obwoe(divergence_method = "kl"))
direct_dmiv <- ob_numerical_dmiv(target = gc_data$target,
                                 feature = gc_data$duration,
                                 divergence_method = "kl")

c(`through obwoe()` = via_obwoe$results$duration$divergence_method,
  `direct wrapper`  = direct_dmiv$divergence_method)
```

The request is dropped without a warning. **If you need an algorithm-specific
parameter, call the engine directly.** The table in each reference entry below
lists which parameters that applies to; an algorithm with no own parameters is
fully configurable through `obwoe()`.

## Coverage

```{r coverage}
coverage <- alg
coverage$feature_types <- ifelse(coverage$numerical & coverage$categorical, "both",
                     ifelse(coverage$numerical, "numerical", "categorical"))
coverage$target <- ifelse(coverage$multinomial, "binary + multinomial", "binary")
coverage[order(coverage$feature_types, coverage$algorithm),
         c("algorithm", "feature_types", "target")]
```

`jedi_mwoe` is the only engine that accepts a multinomial target. With more than
two classes `obwoe(algorithm = "auto")` resolves to it automatically; every other
algorithm requires a binary 0/1 target.

## Choosing an algorithm

`algorithm = "auto"` resolves to `jedi` for a binary target and `jedi_mwoe` for a
multinomial one. It is a fixed alias, not a search: it does not try several
engines and pick a winner.

That default is a reasonable one. On German Credit with five-fold held-out IV,
`jedi` is within one standard error of the best available engine on 18 of 20
variables — so for most variables the choice of algorithm is worth close to
nothing, and the effort is better spent elsewhere. Where it does matter, it
tends to matter on continuous variables with a long tail.

Start from the constraint that actually binds:

| If you need | Reach for |
|---|---|
| A sound general default | `jedi` |
| A multinomial target | `jedi_mwoe` (the only option) |
| Monotonic WoE as a hard requirement | `mob`, `ir`, `mrblp`, `mba`, `gmb` |
| The bin count chosen for you | `mdlp`, `fast_mdlp` |
| A statistical stopping rule | `cm` (chi-square), `fetb` (Fisher exact) |
| The IV-optimal partition under constraints | `dp`, `ivb` |
| High-cardinality categoricals | `sblp`, `swb`, `sketch` |
| Streaming or out-of-core data | `sketch` |
| An unsupervised baseline | `ewb`, `kmb`, `ubsd` |

## Reference

Each entry gives what the engine does, the function to call it directly, the
feature types it accepts, and its own parameters — the ones `obwoe()` cannot
reach.

### Information-theoretic

These optimise entropy, information gain, or a divergence between the class
distributions.

**`jedi`** — Joint Entropy-Driven Interval discretization. Quantile
initialisation followed by merging that balances entropy reduction against
statistical stability. The package default and a sound general-purpose choice.
`ob_numerical_jedi()`, `ob_categorical_jedi()`. Both types. No own parameters.

**`jedi_mwoe`** — the JEDI framework extended to multinomial Weight of Evidence.
The only engine that accepts a target with more than two classes.
`ob_numerical_jedi_mwoe()`, `ob_categorical_jedi_mwoe()`. Both types. No own
parameters. Its `woe` and `iv` are matrices with one column per class, not
vectors — see the caveats section.

**`mdlp`** — Minimum Description Length Principle. Balances the cost of
describing the model against the cost of describing the data, so it chooses the
number of bins itself rather than taking it from you. `ob_numerical_mdlp()`.
Numerical. Own parameter: `laplace_smoothing`.

**`fast_mdlp`** — MDLP with optional WoE monotonicity constraints, optimised for
speed. `ob_numerical_fast_mdlp()`. Numerical. Own parameter:
`force_monotonicity`.

**`dmiv`** — divergence-based hierarchical merging following Zeng (2013).
Maximises a chosen divergence between the class distributions rather than IV.
`ob_numerical_dmiv()`, `ob_categorical_dmiv()`. Both types. Own parameters:
`divergence_method` (`"l2"`, `"kl"`, `"js"`, `"he"`, and others),
`bin_method`, and `is_monotonic` on the numerical side. **Its default
`bin_method = "woe1"` is not standard WoE** — see the caveats section.

**`ivb`** — dynamic programming that maximises total IV exactly, with Bayesian
smoothing for stability. `ob_categorical_ivb()`. Categorical. No own parameters.

**`udt`** — information gain identifies the most informative split points, then
bottom-up merging satisfies the frequency, monotonicity and bin-count
constraints. Despite being called a "User-Defined Technique", and historically
"Unsupervised Decision Tree", it is fully supervised. `ob_numerical_udt()`,
`ob_categorical_udt()`. Both types. Own parameters on the numerical side:
`laplace_smoothing`, `monotonicity_direction`.

### Statistical merging

These merge neighbouring bins until a hypothesis test says the remaining ones
are meaningfully different.

**`cm`** — ChiMerge (Kerber, 1992), with the optional Chi2 extension (Liu &
Setiono, 1995). Merges adjacent bins whose chi-square statistic indicates no
significant difference. `ob_numerical_cm()`, `ob_categorical_cm()`. Both types.
Own parameters: `chi_merge_threshold`, `use_chi2_algorithm`, plus `init_method`
on the numerical side. Note that `obwoe()` cannot reach `use_chi2_algorithm`, so
through the main interface `cm` is always ChiMerge.

**`fetb`** — Fisher's Exact Test as the merge criterion, merging the pair with
the highest p-value. Exact rather than asymptotic, so it stays valid in sparse
bins where chi-square does not. `ob_numerical_fetb()`, `ob_categorical_fetb()`.
Both types. No own parameters.

**`mob`** — Monotonic Optimal Binning. Enforces strictly monotonic WoE, which is
a regulatory requirement in several credit-scoring contexts.
`ob_numerical_mob()`, `ob_categorical_mob()`. Both types. Own parameter on the
numerical side: `laplace_smoothing`.

### Shape-constrained

These treat monotonicity as the primary constraint.

**`ir`** — Isotonic regression by the Pool Adjacent Violators Algorithm.
Guarantees a monotonic relationship between bin index and event rate.
`ob_numerical_ir()`. Numerical. Own parameter: `auto_monotonicity`.

**`mrblp`** — greedy monotonic binning with majority-vote direction detection.
The package's own documentation notes that, despite "Likelihood Ratio
Pre-binning" in the name, it uses equal-frequency pre-binning.
`ob_numerical_mrblp()`. Numerical. Own parameter: `laplace_smoothing`.

**`mblp`** — greedy monotonic binning with guaranteed WoE monotonicity. Despite
the "Linear Programming" in its name it is an iterative heuristic over quantile
pre-bins, not an LP. `ob_numerical_mblp()`. Numerical. Own parameter:
`force_monotonic_direction`.

**`oslp`** — quantile pre-binning plus monotonicity enforcement. As with `mblp`,
the "LP" in the name does not correspond to any linear programme.
`ob_numerical_oslp()`. Numerical. Own parameter: `laplace_smoothing`.

**`gmb`** — greedy bottom-up merging that maximises total IV, with Bayesian
smoothing. `ob_categorical_gmb()`. Categorical. No own parameters.

**`mba`** — Monotonic Binning Algorithm: enforces strict WoE monotonicity while
optimising IV. `ob_categorical_mba()`. Categorical. No own parameters.

**`swb`** — Sliding Window Binning. Frequency-threshold grouping followed by
iterative optimisation toward monotonicity. `ob_categorical_swb()`. Categorical.
No own parameters.

### Combinatorial and dynamic programming

**`dp`** — dynamic programming over the ordered bins, maximising total IV under
the bin-count constraint, with optional monotonic trend enforcement.
`ob_numerical_dp()`, `ob_categorical_dp()`. Both types. Own parameter:
`monotonic_trend` (`"auto"`, `"ascending"`, `"descending"`, `"none"`).

**`bb`** — a branch-and-bound-style search over candidate partitions.
`ob_numerical_bb()`. Numerical. Own parameter: `is_monotonic`.

**`milp`** — despite the name, this is a heuristic merge that maximises IV under
a monotonicity constraint; it does not build or solve a mixed-integer programme.
The package's own documentation says so. `ob_categorical_milp()`. Categorical.
No own parameters.

**`sblp`** — Similarity-Based Logistic Partitioning. Sorts categories by event
rate, then partitions by dynamic programming. Well suited to high-cardinality
categoricals, where sorting by event rate imposes a meaningful order on levels
that have none. `ob_categorical_sblp()`. Categorical. Own parameter: `alpha`.

### Density-based

**`ldb`** — Local Density Binning. Kernel density estimation locates natural
low-density regions to place cut points. `ob_numerical_ldb()`. Numerical. Own
parameter: `enforce_monotonic`.

**`lpdb`** — density estimation plus information-theoretic optimisation, placing
boundaries at critical points of the estimated density curve.
`ob_numerical_lpdb()`. Numerical. Own parameters: `polynomial_degree`,
`enforce_monotonic`.

### Unsupervised initialisation, supervised refinement

These start from the feature's own distribution and then optimise against the
target. They are useful as baselines and when the target is very sparse.

**`ewb`** — equal-width initialisation followed by supervised IV optimisation.
`ob_numerical_ewb()`. Numerical. Own parameter: `is_monotonic`.

**`kmb`** — k-means-inspired initialisation: centroids placed across the range,
cuts at the midpoints, then supervised refinement. `ob_numerical_kmb()`.
Numerical. Own parameter: `enforce_monotonic`.

**`ubsd`** — initialises from the mean and standard deviation, then refines with
WoE and IV. `ob_numerical_ubsd()`. Numerical. Own parameter: `laplace_smoothing`.

### Streaming

**`sketch`** — the only engine built on a fundamentally different computational
model. The numerical variant uses a KLL sketch (Karnin, Lang & Liberty, 2016)
for approximate quantiles; the categorical variant uses a Count-Min sketch for
category frequencies. Both trade exactness for bounded memory, which is what
makes them the right choice for streaming or out-of-core data.
`ob_numerical_sketch()`, `ob_categorical_sketch()`. Both types. Own parameters:
`sketch_k` and `monotonic` on the numerical side, `sketch_width` and
`sketch_depth` on the categorical side.

### Metaheuristic

**`sab`** — simulated annealing over category-to-bin assignments, maximising IV
under a monotonicity constraint. `ob_categorical_sab()`. Categorical. Own
parameters: `initial_temperature`, `cooling_rate`, `adaptive_cooling`. Its
search is seeded from R's RNG stream, so `set.seed()` makes it reproducible.

## Measured behaviour

All 37 combinations on the bundled German Credit data, one numerical feature and
one categorical feature, with identical constraints:

```{r bench, results = "asis"}
run_all <- function(feature, type) {
  ids <- alg$algorithm[alg[[type]]]
  out <- lapply(ids, function(a) {
    r <- try(suppressWarnings(suppressMessages(
      obwoe(gc_data, target = "target", feature = feature,
            algorithm = a, min_bins = 2, max_bins = 5))), silent = TRUE)
    if (inherits(r, "try-error") || r$summary$error) {
      data.frame(algorithm = a, bins = NA_integer_, total_iv = NA_real_)
    } else {
      data.frame(algorithm = a, bins = r$summary$n_bins,
                 total_iv = round(r$summary$total_iv, 4))
    }
  })
  do.call(rbind, out)
}

num_res <- run_all("duration", "numerical")
cat_res <- run_all("purpose", "categorical")
knitr::kable(num_res[order(-num_res$total_iv), ], row.names = FALSE,
             caption = "Numerical engines on `duration`")
```

```{r bench-cat, results = "asis"}
knitr::kable(cat_res[order(-cat_res$total_iv), ], row.names = FALSE,
             caption = "Categorical engines on `purpose`")
```

Read those tables as a sanity check, not as a ranking. IV measured on the same
data that produced the bins rewards more bins automatically, and one variable
cannot separate engines that differ by a few thousandths. A defensible
comparison needs held-out data, which the package's cross-validation-free
interface does not do for you.

### Cost

Every engine binned a 100,000-row numerical feature in under a quarter of a
second, single-threaded, as of version 1.13.3:

| Engine | Seconds | Engine | Seconds |
|---|---|---|---|
| `lpdb`, `ldb` | 0.018 | `kmb` | 0.031 |
| `fetb`, `jedi` | 0.022 | `bb`, `cm` | 0.036 |
| `ir` | 0.024 | `mrblp`, `udt` | 0.037 |
| `jedi_mwoe` | 0.029 | `ewb` | 0.040 |
| `mblp` | 0.048 | `sketch` | 0.050 |
| `dp` | 0.055 | `fast_mdlp` | 0.066 |
| `ubsd` | 0.069 | `mdlp` | 0.076 |
| `mob` | 0.077 | `oslp` | 0.172 |
| `dmiv` | 0.231 | | |

Before 1.13.3 this table looked very different: `ldb`, `lpdb` and numerical
`udt` scaled quadratically and took 10.7s, 10.7s and 7.4s at just 50,000 rows.
They are now among the fastest engines in the package.

## Cross-cutting caveats

**Names do not always describe implementations.** `milp` is not a mixed-integer
programme, `mblp` and `oslp` are not linear programmes, `mrblp` does not use
likelihood-ratio pre-binning, and `udt` is not unsupervised. Each of these is
documented on the algorithm's own help page; they are collected here so the
family names in this vignette are not read as promises.

**`dmiv` reports a non-standard WoE by default.** Its `bin_method` defaults to
`"woe1"`, which is Zeng's log-odds `log((pos + 0.5)/(neg + 0.5))` rather than
`log((pos_i/TP)/(neg_i/TN))`. The two differ by the constant `log(TP/TN)`, so IV
is unaffected — the offset cancels in the sum — but the WoE values are not
comparable with those from the other 27 engines, and a model fitted on them will
have a different intercept. Pass `bin_method = "woe"` to the wrapper directly if
you need the standard definition.

**`jedi_mwoe` returns matrices, not vectors.** With a multinomial target, `woe`
and `iv` carry one column per class. `summary$total_iv` sums across all classes,
which is a number that has no counterpart in the binary case. `obwoe_apply()`
refuses multinomial models rather than silently applying one class's WoE.

**Stochastic engines are reproducible.** `sab`, `kmb` and `mba` involve random
choices, but their generators are seeded from R's own RNG stream, so
`set.seed()` fixes their output exactly.

**`max_bins` is honoured by every engine as of 1.13.3.** Before that release,
`mdlp`, `gmb` and `fetb` could silently exceed it — `mdlp` returned 18 bins
against a cap of 5.

**Pre-binning resolution is a modelling decision.** `max_n_prebins` runs before
any of these engines sees the data, and its default of 20 can smear a heavy tail
before optimisation begins. See `?control.obwoe`, which carries the measurements
and the reason the default is nonetheless left alone.

## References

Kerber, R. (1992). ChiMerge: Discretization of numeric attributes. *AAAI-92*.

Liu, H. & Setiono, R. (1995). Chi2: Feature selection and discretization of
numeric attributes. *ICTAI-95*.

Fayyad, U. & Irani, K. (1993). Multi-interval discretization of continuous-valued
attributes for classification learning. *IJCAI-93*.

Karnin, Z., Lang, K. & Liberty, E. (2016). Optimal quantile approximation in
streams. *FOCS 2016*.

Zeng, G. (2013). Metric divergence measures and information value in credit
scoring. *Journal of Mathematics*.

Siddiqi, N. (2006). *Credit Risk Scorecards*. Wiley.
